A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-02 latex/main.tex
Fixup references
Updates from Overleaf

[TODO revise this paragraph to give some background info, and reference
the Viridian work]
For Omicron BA.1 and BA.2, we used SRR17041376 and SRR17461792, respectively.
We chose SRR17041376 (a sample dated 2021-11-12 from South Africa)
as the BA.1 seed (https://github.com/jeromekelleher/sc2ts-paper/issues/264),
as it could be well explained as as a non-recombinant given the ARG available on its collection date.
For BA.2, we chose SRR17461792 (dated 2021-11-27 from South Africa),
which is the earliest BA.2 sample present in the Viridian v04 dataset.
It is dated later than SRR17041376 (the BA.1 seed),
allowing time for the BA.1 lineage to accumulate enough sampled descendants
and to get established in the ARG.
See Figure~\ref{fig:subgraph_ba1_ba2} for an overview of the resulting 
subgraph.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed
a more comprehensive mutational landscape of SARS-CoV-2
by mapping deletions onto the ARG using post-hoc parsimony.
We identified major deletion events, for example,
the deletions inherited by at least 10,000 samples
shown in Table~\ref{tab:major_dels}.
% This just says that the deletions are found in sequences of VOCs.
All these deletions are known to occur in major VOCs 
(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
\url{https://github.com/cov-lineages/pango-designation/issues/361};
\cite{Stern2021,Tegally2022,Li2023indels}),
except the 4-base deletion and two 5-base deletions in ORF1ab,
which are likely artefacts of deletion remapping or
sequence misalignment (see below).
Notably, eight of these deletions occur
in the N-terminal domain of the S1 unit in Spike,
a known hotspot for recurrent deletions \cite{McCarthy2021}
(also see Table~\ref{tab:rec_dels}).

% Placement of deletion events linked to VOCs
In the ARG, we found that the following deletion events
are associated with the origins of major VOCs
(Table~\ref{tab:major_dels}):
a 9-base deletion at 11288,
a 6-base deletion at 21,765, and
a 3-base deletion at 21,991 above the Alpha origin node;
a 6-base deletion at 22,029 above the Delta origin node;
a 6-base deletion at 28,248 above a child of the Delta origin node;
a 9-base deletion at 28,362 above the B.1.1.529 origin node;
a 3-base deletion at 6,513,
a 3-base deletion at 22,194,
a 6-base deletion at 21,765, and
a 9-base deletion at 21,987 above the BA.1 origin node;
a 9-base deletion at 21,633 above the BA.2 origin node; and
a 6-base deletion at 21,633 above the BA.4 origin node.
All these deletion events have been proposed previously
to be linked with the emergence of these VOCs
(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
\url{https://github.com/cov-lineages/pango-designation/issues/361};
\cite{Stern2021,Tegally2022}).

% Known recurrent deletions
Using the ARG, we identified highly recurrent deletions,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG
shown in Table~\ref{tab:rec_dels}.
These include known recurrent deletions, for example,
a 6-base deletion at 21,765,
a 3-base deletion at 21,991, and
a 3-base deletion 22,194 in S1-NTD
associated with antibody escape \cite{McCarthy2021,Meng2021},
which are also caused by major deletion events in the ARG
(Table~\ref{tab:major_dels}).

% Problems and future improvement
Detection of deletions and inference of deletion events
are difficult, especially when the dataset is enormous.
Here we used a naive procedure to map deletions onto the ARG,
treating each site involved in a deletion
as independent of the other sites.
Consequently, some deletion events might be split up and
appear as multiple smaller deletion events,
which are likely artefactual.
The three contiguous out-of-frame deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292)
are examples of such artefacts (Table~\ref{tab:major_dels}).
In the sequence alignments, there are two 9-base deletions,
one at 11,283 which commonly occur in BA.1 samples and
the other at 11,288 which commonly occur in BA.2 samples.
Assuming that the alignments are correct,
there should be two deletion events,
but they were mapped as three deletion events
because they had four overlapping sites from 11,288 to 11,292,
which were inferred by parsimony
as the 4-base deletion event at 11,288.
Furthermore, sequencing errors or sequence misalignments
can also cause artefactual deletions,
for example, the out-of-frame recurrent deletions
in N, ORF1ab, ORF8, and particularly S1-NTD
(e.g., the 2-base deletion at 21,765)
(Table~\ref{tab:rec_dels}).
Future improvements in deletion remapping and data pre-processing
should remove such artefactual deletions.

\begin{table}
\centering
\caption{
Major deletion events in the ARG.
These events are inherited by
at least 10,000 samples in the ARG.
Count is the number of samples inheriting a deletion.
The Pango lineage assigned to the node
above which an event occurs is shown.
Abbreviations: NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:major_dels}
\begin{tabular}{rrcrcrr}
\toprule
Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 871782 & 35.1220 \\
28362 & 9 & N & 1436802 & B.1.1.529 (Omicron) & 871774 & 35.1216 \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 \\
\bottomrule
\end{tabular}
\end{table}

>>>>

[TODO revise this paragraph to give some background info, and reference
the Viridian work]
For Omicron BA.1 and BA.2, we used SRR17041376 and SRR17461792, respectively.
We chose SRR17041376 (a sample dated 2021-11-12 from South Africa)
as the BA.1 seed (https://github.com/jeromekelleher/sc2ts-paper/issues/264),
as it could be well explained as as a non-recombinant given the ARG available on its collection date.
For BA.2, we chose SRR17461792 (dated 2021-11-27 from South Africa),
which is the earliest BA.2 sample present in the Viridian v04 dataset.
It is dated later than SRR17041376 (the BA.1 seed),
allowing time for the BA.1 lineage to accumulate enough sampled descendants
and to get established in the ARG.
See Figure~\ref{fig:subgraph_omicron} for an overview of the resulting 
subgraph.

@@

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
We reconstructed
a more comprehensive mutational landscape of SARS-CoV-2
by mapping deletions onto the ARG using post-hoc parsimony.
We identified major deletion events, for example,
the deletions inherited by at least 10,000 samples
shown in Table~\ref{tab:major_dels}.
% This just says that the deletions are found in sequences of VOCs.
All these deletions are known to occur in major VOCs 
(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
\url{https://github.com/cov-lineages/pango-designation/issues/361};
\cite{Stern2021,Tegally2022,Li2023indels}),
except the 4-base deletion and two 5-base deletions in ORF1ab,
which are likely artefacts of
deletion remapping or sequence misalignment (see below).
Notably, eight of these deletions occur
in the N-terminal domain of the S1 unit in Spike,
a known hotspot for recurrent deletions \cite{McCarthy2021}
(also see Table~\ref{tab:rec_dels}).
In the ARG, we found that most of the major deletion events
are associated with the origins of major VOCs
(Table~\ref{tab:major_dels}).
They have been linked previously with the emergence of these VOCs
(\url{https://virological.org/t/preliminary-genomic-characterisation-of-an-emergent-sars-cov-2-lineage-in-the-uk-defined-by-a-novel-set-of-spike-mutations/563};
\url{https://github.com/cov-lineages/pango-designation/issues/361};
\cite{Stern2021,Tegally2022}).

Using the ARG, we identified highly recurrent deletions,
for example, the top 20 deletions in terms of
the number of occurrences in the ARG
shown in Table~\ref{tab:rec_dels}.
These include known recurrent deletions, for example,
a 6-base deletion at 21,765,
a 3-base deletion at 21,991, and
a 3-base deletion 22,194 in S1-NTD
associated with antibody escape \cite{McCarthy2021,Meng2021},
which are also caused by major deletion events in the ARG
(Table~\ref{tab:major_dels}).

Detection of deletions and inference of deletion events
are difficult, especially when the dataset is enormous.
Here we used a naive procedure to map deletions onto the ARG,
treating each site involved in a deletion
as independent of the other sites.
Consequently, some deletion events might be split up and
appear as multiple smaller deletion events,
which are likely artefactual.
The three contiguous out-of-frame deletions in NSP6
(5-base at 11,283, 4-base at 11,288, and 5-base at 11,292)
are examples of such artefacts (Table~\ref{tab:major_dels}).
In the sequence alignments, there are two 9-base deletions,
one at 11,283 which commonly occur in BA.1 samples and
the other at 11,288 which commonly occur in BA.2 samples.
Assuming that the alignments are correct,
there should be two deletion events,
but they were mapped as three deletion events
because they had four overlapping sites from 11,288 to 11,292,
which were inferred by parsimony
as the 4-base deletion event at 11,288.
Furthermore, sequencing errors or sequence misalignments
can also cause artefactual deletions,
for example, the out-of-frame recurrent deletions
in N, ORF1ab, ORF8, and particularly S1-NTD
(e.g., the 2-base deletion at 21,765)
(Table~\ref{tab:rec_dels}).
Future improvements in deletion remapping and data pre-processing
should remove such artefactual deletions.

\begin{table}
\centering
\caption{
Major deletion events in the ARG.
These events are inherited by
at least 10,000 samples in the ARG.
Count is the number of samples inheriting a deletion.
The Pango lineage assigned to the node
above which an event occurs is shown.
Abbreviations: NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\begin{tabular}{rrcrcrrc}
\toprule
Start & Length & Region & Node & Pango lineage & Count & Frequency (\%) & Ref\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 &  \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 & B \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 & B \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 871782 & 35.1220 &  \\
28362 & 9 & N & 1436802 & B.1.1.529 (Omicron) & 871774 & 35.1216 & C \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 & D \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 &  \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 & D \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 &  \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 & D \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 & D \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 & D \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 &  \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 & A \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 & A \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 & A \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 & E \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 &  \\
\bottomrule
\end{tabular}
\end{table}
